When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2017-10-27 paper/paper.tex
Missing author
More things

\usepackage{showframe}

@@

\jairheading{1}{2016}{0-0}{0/0}{0/0}
\ShortHeadings{When Subgraph Isomorphism is Really Hard}
{McCreesh, Prosser \& Trimble}
\firstpageno{1}

@@

\crefname{algorithm}{Algorithm}{Algorithms}
\Crefname{algorithm}{Algorithm}{Algorithms}
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{table}{Table}{Tables}
\Crefname{table}{Table}{Tables}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}
\crefname{proposition}{Proposition}{Propositions}
\Crefname{proposition}{Proposition}{Propositions}

@@

\definecolor{uofgsandstone}{rgb}{0.321569, 0.278431, 0.231373}
\definecolor{uofglawn}{rgb}{0.517647, 0.741176, 0}
\definecolor{uofgcobalt}{rgb}{0, 0.615686, 0.92549}
\definecolor{uofgpumpkin}{rgb}{1.0, 0.72549, 0.282353}
\definecolor{uofgthistle}{rgb}{0.584314, 0.070588, 0.447059}

@@

\title{When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph
Databases\thanks{Part of this paper appeared at IJCAI \citep{DBLP:conf/ijcai/McCreeshPT16};
\cref{section:introduction} has been expanded and
\cref{section:models,section:labelled,section:filterverify} are new material.}}

@@

\begin{abstract}
    The subgraph isomorphism problem involves finding a copy of a pattern graph inside a larger
    target graph. The non-induced version allows extra edges in the target, whilst the induced
    version does not.  Although both subgraph isomorphism variants are \NP-complete, modern
    algorithms can operate comfortably on most real-world graphs with thousands of vertices.
    However, they cannot handle arbitrary instances of this size. We show how to generate ``really
    hard'' random instances for subgraph isomorphism problems, which are computationally challenging
    with only a couple of hundred vertices in the target, and only twenty to thirty pattern
    vertices. For the non-induced version of the problem, these instances lie on a satisfiable /
    unsatisfiable phase transition, whose location we can predict; for the induced variant, much
    richer behaviour is observed, and constrainedness gives a better measure of difficulty than does
    proximity to a phase transition. In both cases, we can use this to understand and improve
    variable and value ordering heuristics.

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a given pattern graph to the vertices of a given target graph which preserves
adjacency---in essence, we are ``finding a copy of'' the pattern inside the target. The
\emph{induced} variant of the problem additionally requires that the mapping preserve non-adjacency,
so there are no ``extra edges'' in the copy of the pattern that we find. We illustrate both variants
in \cref{figure:sip}.  Although these problems are \NP-complete, modern practical subgraph
isomorphism algorithms can handle problem instances with many hundreds of vertices in the pattern
graph, and up to ten thousand vertices in the target graph
\citep{DBLP:journals/pami/CordellaFSV04,DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16},
leading to successful applications in areas such as computer vision
\citep{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, biochemistry
\citep{o:10.1371/journal.pone.0076911,DBLP:conf/gbrpr/CarlettiFV15}, and pattern recognition
\citep{DBLP:journals/ijprai/ConteFSV04}.  In particular, a labelled version of subgraph isomorphism
is one of the key components in supporting complex queries in graph databases---typically, these
systems store a fixed set of target graphs, and for a sequence of pattern queries, they must return
every target graph which contains that pattern.  One common approach to this problem is to combine a
subgraph isomorphism algorithm with an indexing system in a so-called ``filter~/ verify'' model,
where invariants are used to attempt to pre-exclude unsatisfiable instances to avoid the cost of a
subgraph isomorphism call. In this context, the terms \emph{matching} and \emph{verification} are
often used for the subgraph isomorphism step (or a slightly broader problem, for example permitting
wildcards).

\begin{figure}[t]
    \centering
    \begin{tikzpicture}[scale=0.3,every node/.style={font=\footnotesize}]
        \begin{scope}
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Na) at (1,  0) {\phantom{0}};
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Nb) at (1, -2) {\phantom{0}};
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Nc) at (0, -4) {\phantom{0}};
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Nd) at (2, -4) {\phantom{0}};

@@

The first contribution of this paper is to present and evaluate new methods for creating random
pattern~/ target pairs.  The method we introduce generates both satisfiable and unsatisfiable
instances, and can produce computationally challenging instances with only a few tens of vertices in
the pattern, and 150 vertices in the target. Note that the lack of
unsatisfiable instances for testing purposes cannot be addressed simply by taking a pattern graph
from one of the existing random suites with the ``wrong'' target graph, as this tends to give either
a trivially unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the
case that a relatively small random graph is unlikely to appear in a larger random graph.)

@@

For certain choices of parameters for induced isomorphisms, there are two phase transitions, going
from satisfiable to unsatisfiable, and then from unsatisfiable back to satisfiable. Again, when
going from satisfiable to unsatisfiable (from either direction), instances go from being trivial to
really hard to solve. However, each of the three solvers we test also finds the central
unsatisfiable region to be hard, despite it not being near a phase transition. To show that this is
not a simple weakness of current subgraph isomorphism algorithms, we verify that this region is also
hard for boolean satisfiability, pseudo-boolean, and mixed integer solvers, and under reduction to
the clique problem. Interestingly, the constrainedness measure proposed by
\citet{DBLP:conf/aaai/GentMPW96} \emph{does} predict this difficult region---the third contribution
of this paper is to use these instances to provide evidence in favour of constrainedness, rather
than proximity to a phase transition, being an accurate predictor of difficulty, and to show that
constrainedness is not simply a refinement of a phase transition prediction.

% What about other random graph models? Graphs where every vertex has the same degree, which are known
% as $k$-regular, exhibit very similar characteristics.

When labels on vertices are introduced, as is commonly seen in real-world applications and in graph
database systems, richer behaviour emerges, particularly when moving away from a uniform labelling
scheme, and we see that the popular VF2 algorithm behaves much worse than two more recent
constraint-based algorithms in certain cases.  A close look shows that these cases \emph{should} be
easy to solve.

@@

We look at some of the datasets commonly used to test graph database systems, which the literature
suggests are hard to solve. Simple experiments show that the entire perceived difficulty
of each of these datasets comes down to the widespread use of especially poor subgraph isomorphism
algorithms. Our fourth contribution is to show that the use of forward-checking and the right
variable ordering heuristic, even without the more sophisticated recent advances in subgraph
algorithms, makes the entire graph database filter~/ verify paradigm unnecessary.  Finally, we
explain why filter~/ verify (and other recently proposed techniques) cannot be beneficial even on
more challenging instances except if an extremely poor choice of subgraph isomorphism algorithm is
made.

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops.  The \emph{order} of a
graph is the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the vertex set of a
graph $G$. The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the graph with the
same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and only if $v$
and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with $n$
vertices, and an edge between each distinct pair of vertices with independent probability $p$.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. An \emph{induced subgraph
isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$,
then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P \rightarrowtail T$
for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced isomorphism. Observe
that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced isomorphism $i : P
\rightarrowtail T$ which is also a non-induced isomorphism $\shortoverline{i} : \shortoverline{P}
\rightarrowtail \shortoverline{T}$.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs, running
Scientific Linux release 6.7. We will be working with three subgraph isomorphism solvers: the
Glasgow solver \citet{DBLP:conf/cp/McCreeshP15}; LAD \citep{DBLP:journals/ai/Solnon10}; and VF2
\citep{DBLP:journals/pami/CordellaFSV04}. Each was compiled using GCC 4.9.

The Glasgow and LAD solvers store domains, and use backtracking search to build up an assignment of
target vertices (values) to pattern vertices (variables), but differ in terms of inference and
ordering heuristics.  The approach used by VF2 is similar, although the domains of variables are not
stored (in the style of conventional backtracking, rather than forward-checking), and so domain
wipeouts are not detected until an assignment is made. Both the Glasgow and LAD solvers use domain
sizes to guide search, whilst VF2 uses adjacency to previously-assigned vertices.

We measure only the number of recursive calls (guessed assignments) made, not
runtimes. We are not aiming to compare absolute performance between solvers; rather, we are looking
for solver-independent patterns of difficulty. All experiments use a timeout of 1,000 seconds, which
is enough for the Glasgow solver to solve nearly all instances (whose orders were selected with this
timeout in mind), although we may slightly overestimate the proportion of unsatisfiable instances
for extremely sparse or dense pattern graphs. The LAD and VF2 solvers experienced many more failures
with this timeout, so our picture of just how hard the hardest instances are with these solvers is
less detailed.

?? Glasgow not parallel

?? VF3 does not do non-induced

\section{Non-Induced Subgraph Isomorphisms}

@@

\begin{figure}[tb]
    \centering
    \begin{tikzpicture}[every node/.style={inner sep=0pt, outer sep=0pt}]
        \matrix {
            \node [anchor=center] {}; &
            \node [anchor=center] {\scriptsize $G(10, x)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(20, x)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(30, x)$\hspace*{3em}}; &
            \\
            \node [anchor=center, inner sep=4pt] [rotate=90] {\scriptsize Satisfiable?}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-satisfiable-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-satisfiable-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-satisfiable-30-150.pdf}}; &
            \\[0.3cm]
            \node [anchor=center, inner sep=4pt] [rotate=90] {\scriptsize Glasgow}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-nodes-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-nodes-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-nodes-30-150.pdf}}; &
            \\[0.3cm]
            \node [anchor=center, inner sep=4pt] [rotate=90] {\scriptsize LAD}; &
            \node [anchor=center] {\includegraphics*{plots/graph-lad-non-induced-nodes-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-lad-non-induced-nodes-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-lad-non-induced-nodes-30-150.pdf}}; &
            \\[0.3cm]
            \node [anchor=center, inner sep=4pt] [rotate=90] {\scriptsize VF2}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-nodes-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-nodes-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-nodes-30-150.pdf}}; \\
        };
    \end{tikzpicture}

    \caption{Behaviour of algorithms on the non-induced variant. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands show the phase transitions, and the black lines are the predictions using
        equation~\eqref{equation:non-induced-prediction} of where the
        phase transition will occur. On the subsequent three rows, we show the number of search nodes used by the
        Glasgow, LAD and VF2 solvers; the dark regions indicate ``really hard'' instances, and black
        points indicate that at least one timeout occurred.}
    \label{figure:non-induced}
\end{figure}

@@

On subsequent rows, we show the average number of search nodes used by the different algorithms. In
general, satisfiable instances are easy, until very close to the phase transition. As we hit the
phase transition and move into the unsatisfiable region, we see complexity increase. Finally, as we
pass through the phase transition and move deeper into the unsatisfiable region, instances become
easier again. This behaviour is largely solver-independent, although VF2 has a larger hard region
than Glasgow or LAD. VF2 also finds some instances extremely hard when the target is empty but the
pattern is not---this turns out to be extremely important, and we return to it in
\cref{section:labelled,section:filterverify}. Thus, although we have moved away from a single
control parameter, we still observe the easy-hard-easy pattern seen in many \NP-complete
problems.

@@

Various general principles have been considered when designing variable and value ordering
heuristics for backtracking search algorithms---one of these is to try to maximise the expected
number of solutions inside any subproblem considered during search \citep{DBLP:conf/cp/GentMPSW96}.  This is
usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD and
Glasgow pick a variable with fewest remaining values in its domain: doing this will generally reduce
the first part of the $\langle Sol \rangle$ equation by as little as possible. When two or more
domains are of equal size, LAD simply breaks ties lexicographically, whereas Glasgow will pick a
variable corresponding to a pattern vertex of highest degree. This strategy was determined
empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking a pattern
vertex of high degree will make the remaining pattern subgraph sparser, which will decrease the
exponent in the second half of the formula, maximising the overall value. LAD does not apply a value
ordering heuristic, but Glasgow does: it prefers target vertices of highest degree.  Again, this was
determined empirically, but it has the effect of increasing $\langle Sol \rangle$ by leaving as many
vertices as possible available for future use. The VF2 heuristics, in contrast, are based around
preserving connectivity, which gives very little discrimination except on the sparsest of inputs.

\section{Induced Subgraph Isomorphisms}

?? LAD and Glasgow not really designed for induced

@@

            \node [anchor=center] {\scriptsize $G(10, x)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(14, x)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(15, x)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(16, x)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(20, x)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(30, x)$\hspace*{3em}}; &

@@

            \node [anchor=center, inner sep=4pt, rotate=90] {\scriptsize Satisfiable?}; &

@@

            \node [anchor=center, inner sep=4pt, rotate=90] {\scriptsize Glasgow}; &

@@

            \node [anchor=center, inner sep=4pt, rotate=90] {\scriptsize LAD}; &

@@

            \node [anchor=center, inner sep=4pt, rotate=90] {\scriptsize VF2}; &

@@

            \node [anchor=center, inner sep=4pt, rotate=90] {\scriptsize VF3}; &

@@

            \node [anchor=center, inner sep=4pt, rotate=90] {\scriptsize Constrainedness}; &

@@

            \node [anchor=center, inner sep=4pt, rotate=90] {\scriptsize Complement?}; &

@@

    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The sixth row plots constrainedness using
    equation~\eqref{equation:constrainedness}: the
    darkest region is where $\kappa = 1$, and the lighter regions show where the problem is either
    over- or under-constrained. The final row shows when the Glasgow algorithm performs better when
    given the complements of the pattern and target graphs as inputs---the solid lines show the
    location of the phase transition, and the dotted lines are $t_d=0.5$ and the $p_d=t_d$
    diagonal.}\label{figure:induced}

@@

            \node [anchor=center] {\scriptsize $G(10, x)$\hspace*{1em}}; &

>>>>

% \usepackage{showframe}

@@

\jairheading{1}{2017}{0-0}{0/0}{0/0}
\ShortHeadings{When Subgraph Isomorphism is Really Hard}
{McCreesh, Prosser, Solnon \& Trimble}
\firstpageno{1}

@@

\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}

@@

@@

\title{When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph
Databases}

@@

\begin{abstract}
    The subgraph isomorphism problem involves finding a copy of a pattern graph inside a larger
    target graph. The non-induced version allows extra edges in the target, whilst the induced
    version does not.  Although both subgraph isomorphism variants are \NP-complete, modern
    algorithms can operate comfortably on most real-world graphs with thousands of vertices.
    However, they cannot handle arbitrary instances of this size. We show how to generate ``really
    hard'' random instances for subgraph isomorphism problems, which are computationally challenging
    with only a couple of hundred vertices in the target, and only twenty to thirty pattern
    vertices. For the non-induced version of the problem, these instances lie on a satisfiable /
    unsatisfiable phase transition, whose location we can predict; for the induced variant, much
    richer behaviour is observed, and constrainedness gives a better measure of difficulty than does
    proximity to a phase transition. In both cases, we can use this to understand and improve
    variable- and value-ordering heuristics.

@@

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a given pattern graph to the vertices of a given target graph which preserves
adjacency---in essence, we are ``finding a copy of'' the pattern inside the target. The
\emph{induced} variant of the problem additionally requires that the mapping preserve non-adjacency,
so there are no ``extra edges'' in the copy of the pattern that we find. We illustrate both variants
in \cref{figure:sip}.  Although these problems are \NP-complete, modern practical subgraph
isomorphism algorithms can handle problem instances with many hundreds of vertices in the pattern
graph, and up to ten thousand vertices in the target graph
\citep{DBLP:journals/pami/CordellaFSV04,DBLP:journals/ai/Solnon10,DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16},
leading to successful applications in areas such as computer vision
\citep{DBLP:journals/cviu/DamiandSHJS11,DBLP:journals/pr/SolnonDHJ15}, biochemistry
\citep{o:10.1371/journal.pone.0076911,DBLP:conf/gbrpr/CarlettiFV15}, and pattern recognition
\citep{DBLP:journals/ijprai/ConteFSV04}.  In particular, a labelled version of subgraph isomorphism
is one of the key components in supporting complex queries in graph databases, which we discuss in
detail in \cref{section:filterverify}. Typically, these systems store a fixed set of target graphs,
and for a sequence of pattern queries, they must return every target graph which contains that
pattern.  One common approach to this problem is to combine a subgraph isomorphism algorithm with an
indexing system in a so-called ``filter~/ verify'' model, where invariants are used to attempt to
pre-exclude unsatisfiable instances to avoid the cost of a subgraph isomorphism call. In this
context, the terms \emph{matching} and \emph{verification} are often used for the subgraph
isomorphism step (or a slightly broader problem, for example permitting wildcards).

\begin{figure}[tb]
    \centering
    \begin{tikzpicture}[scale=0.3,every node/.style={font=\footnotesize}]
        \begin{scope}
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Na) at (1,  0) {\phantom{0}};
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Nb) at (1, -2) {\phantom{0}};
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Nc) at (0, -4) {\phantom{0}};
            \node[draw, circle, fill=chromaygb2, inner sep=0.5pt] (Nd) at (2, -4) {\phantom{0}};

@@

The first contribution of this paper is to present and evaluate new methods for creating random
pattern~/ target pairs.  The method we introduce in \cref{section:non-induced} generates both
satisfiable and unsatisfiable instances, and can produce computationally challenging instances with
only a few tens of vertices in the pattern, and 150 vertices in the target. Note that the lack of
unsatisfiable instances for testing purposes cannot be addressed simply by taking a pattern graph
from one of the existing random suites with the ``wrong'' target graph, as this tends to give either
a trivially unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the
case that a relatively small random graph is unlikely to appear in a larger random graph.)

@@

In \cref{section:induced} we look at the induced isomorphism problem.  For certain choices of
parameters, there are two phase transitions, going from satisfiable to unsatisfiable, and then from
unsatisfiable back to satisfiable. Again, when going from satisfiable to unsatisfiable (from either
direction), instances go from being trivial to really hard to solve. However, each of the three
solvers we test also finds the central unsatisfiable region to be hard, despite it not being near a
phase transition. To show that this is not a simple weakness of current subgraph isomorphism
algorithms, we verify that this region is also hard for boolean satisfiability, pseudo-boolean, and
mixed integer solvers, and under reduction to the clique problem. Interestingly, the constrainedness
measure proposed by \citet{DBLP:conf/aaai/GentMPW96} \emph{does} predict this difficult region---the
third contribution of this paper is to use these instances to provide evidence in favour of
constrainedness, rather than proximity to a phase transition, being an accurate predictor of
difficulty, and to show that constrainedness is not simply a refinement of a phase transition
prediction.

What about other random graph models? In \cref{section:regular} we see that graphs where every vertex
has the same degree (known as $k$-regular) exhibit very similar characteristics.  However, when
labels on vertices are introduced (as is commonly seen in real-world applications and in graph
database systems), richer behaviour emerges, particularly when moving away from a uniform labelling
scheme, and the popular VF2 algorithm behaves much worse than two more recent
constraint-based algorithms in certain cases. In \cref{section:labelled} we take a close look at
these cases, and argue that they \emph{should} be easy to solve.

@@

In \cref{section:filterverify} we look at some of the datasets commonly used to test graph database
systems, which the literature suggests are hard to solve. Simple experiments show that the entire
perceived difficulty of each of these datasets comes down to the widespread use of especially poor
subgraph isomorphism algorithms. Our fourth contribution is to show that the use of forward-checking
and the right variable ordering heuristic, even without the more sophisticated recent advances in
subgraph algorithms, makes the entire graph database filter~/ verify paradigm unnecessary.  Finally,
we explain why filter~/ verify (and other recently proposed techniques) cannot be beneficial even on
more challenging instances except if an extremely poor choice of subgraph isomorphism algorithm is
made.

\bigskip

Part of this paper appeared at IJCAI \citep{DBLP:conf/ijcai/McCreeshPT16};
\cref{section:introduction} has been expanded and
\cref{section:regular,section:labelled,section:filterverify} are new material.

@@

Throughout, our graphs are undirected, and do not have any loops. In some cases vertices have
labels, which are treated as integers; we write $\ell(v)$ for the label of vertex $v$. The
\emph{order} of a graph is the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the
vertex set of a graph $G$. The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and
only if $v$ and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random
graph with $n$ vertices, and an edge between each distinct pair of vertices with independent
probability $p$.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. Where labels are present,
the mapping must also preserve labels, so $\ell(v) = \ell(i(v))$ for any $v$. An \emph{induced
subgraph isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not
adjacent in $P$, then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P
\rightarrowtail T$ for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced
isomorphism. Observe that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced
isomorphism $i : P \rightarrowtail T$ which is also a non-induced isomorphism $\shortoverline{i} :
\shortoverline{P} \rightarrowtail \shortoverline{T}$.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs, running
Scientific Linux release 6.7. We will be working with four subgraph isomorphism solvers: the Glasgow
solver \citet{DBLP:conf/cp/McCreeshP15}; LAD \citep{DBLP:journals/ai/Solnon10}; VF2
\citep{DBLP:journals/pami/CordellaFSV04}; and VF3 \citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was
compiled using GCC 4.9.

The Glasgow and LAD solvers store domains, and use backtracking search to build up an assignment of
target vertices (values) to pattern vertices (variables), but differ in terms of inference and
ordering heuristics.  The approach used by VF2 and VF3 is similar, although the domains of variables
are not stored (in the style of conventional backtracking, rather than forward-checking), and so
domain wipeouts are not detected until an assignment is made. Both the Glasgow and LAD solvers use
domain sizes to guide search, whilst VF2 and VF3 use adjacency to previously-assigned vertices.

We measure only the number of recursive calls (guessed assignments) made, not
runtimes. We are not aiming to compare absolute performance between solvers; rather, we are looking
for solver-independent patterns of difficulty. All experiments use a timeout of 1,000 seconds, which
is enough for the Glasgow solver to solve nearly all instances (whose orders were selected with this
timeout in mind), although we may slightly overestimate the proportion of unsatisfiable instances
for extremely sparse or dense pattern graphs. The LAD, VF2, and VF3 solvers experienced many more failures
with this timeout, so our picture of just how hard the hardest instances are with these solvers is
less detailed.

The Glasgow, LAD, and VF2 solvers support both non-induced and induced isomorphisms (although
neither Glasgow nor LAD were designed with the induced version in mind), whilst VF3 supports only
the induced problem. Note also that VF3 will sometimes crash on especially sparse graphs: we have
treated these instances as timeouts.

\section{Non-Induced Subgraph Isomorphisms}\label{section:non-induced}

@@

\begin{figure}[tb]
    \centering
    \begin{tikzpicture}[every node/.style={inner sep=0pt, outer sep=0pt}]
        \matrix {
            \node [anchor=center] {}; &
            \node [anchor=center] {\scriptsize $G(10, x) \rightarrowtail G(150, y)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(20, x) \rightarrowtail G(150, y)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(30, x) \rightarrowtail G(150, y)$\hspace*{3em}}; &
            \\
            \node [anchor=center, inner sep=2pt] [rotate=90] {\scriptsize Satisfiable?}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-satisfiable-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-satisfiable-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-satisfiable-30-150.pdf}}; &
            \\[0.3cm]
            \node [anchor=center, inner sep=2pt] [rotate=90] {\scriptsize Glasgow}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-nodes-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-nodes-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-non-induced-nodes-30-150.pdf}}; &
            \\[0.3cm]
            \node [anchor=center, inner sep=2pt] [rotate=90] {\scriptsize LAD}; &
            \node [anchor=center] {\includegraphics*{plots/graph-lad-non-induced-nodes-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-lad-non-induced-nodes-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-lad-non-induced-nodes-30-150.pdf}}; &
            \\[0.3cm]
            \node [anchor=center, inner sep=2pt] [rotate=90] {\scriptsize VF2}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-nodes-10-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-nodes-20-150.pdf}}; &
            \node [anchor=center] {\includegraphics*{plots/graph-vf2-non-induced-nodes-30-150.pdf}}; \\
        };
    \end{tikzpicture}

    \caption{Behaviour of algorithms on the non-induced variant, using target graphs of 150
        vertices, and pattern graphs of 10, 20, or 30 vertices. For each plot, the x-axis is the
        pattern edge probability and the y-axis is the target edge probability, both from 0 to 1.
        Along the top row, we show the proportion of instances which are satisfiable; the white
        bands show the phase transitions, and the black lines are the predictions using
        equation~\eqref{equation:non-induced-prediction} of where the
        phase transition will occur. On the subsequent three rows, we show the number of search nodes used by the
        Glasgow, LAD and VF2 solvers; the dark regions indicate ``really hard'' instances, and black
        points indicate that at least one timeout occurred.}
    \label{figure:non-induced}
\end{figure}

@@

On subsequent rows, we show the average number of search nodes used by the different algorithms.
Darker regions indicate harder difficulties, and black is used to indicate that at least one timeout
occurred. In general, satisfiable instances are easy, until very close to the phase transition. As
we hit the phase transition and move into the unsatisfiable region, we see complexity increase.
Finally, as we pass through the phase transition and move deeper into the unsatisfiable region,
instances become easier again. This behaviour is largely solver-independent, although VF2 has a
larger hard region than Glasgow or LAD. Thus, although we have moved away from a single control
parameter, we still observe the easy-hard-easy pattern seen in many \NP-complete problems.

Finally, a close look at the very bottom left of the plots shows that VF2 will sometimes give a
timeout on instances where the target is empty but the pattern is not---this turns out to be
extremely important, and we return to it in \cref{section:labelled,section:filterverify}.

@@

Various general principles have been considered when designing variable and value ordering
heuristics for backtracking search algorithms---one of these is to try to maximise the expected
number of solutions inside any subproblem considered during search \citep{DBLP:conf/cp/GentMPSW96}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow pick a variable with fewest remaining values in its domain: doing this will generally
reduce the first part of the $\langle Sol \rangle$ equation by as little as possible. When two or
more domains are of equal size, LAD simply breaks ties lexicographically, whereas Glasgow will pick
a variable corresponding to a pattern vertex of highest degree (?? check carefully!). This strategy
was determined empirically, but could have been derived from the $\langle Sol \rangle$ formula:
picking a pattern vertex of high degree will make the remaining pattern subgraph sparser, which will
decrease the exponent in the second half of the formula, maximising the overall value. LAD does not
apply a value ordering heuristic, but Glasgow does: it prefers target vertices of highest degree (??
check carefully!).  Again, this was determined empirically, but it has the effect of increasing
$\langle Sol \rangle$ by leaving as many vertices as possible available for future use. The VF2
heuristics, in contrast, are based around preserving connectivity, which gives very little
discrimination except on the sparsest of inputs.

\subsection{Instances for Benchmarking}

?? Instances for benchmarking

?? Something about distance-based filtering being no use on these instances

?? Low degree spread

\section{Induced Subgraph Isomorphisms}\label{section:induced}

@@

            \node [anchor=center] {\scriptsize $G(10, x) \hookrightarrow$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(14, x) \hookrightarrow$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(15, x) \hookrightarrow$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(16, x) \hookrightarrow$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(20, x) \hookrightarrow$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(30, x) \hookrightarrow$\hspace*{3em}}; &

@@

            \node [anchor=center] {\scriptsize $G(150, y)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(150, y)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(150, y)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(150, y)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(150, y)$\hspace*{1em}}; &
            \node [anchor=center] {\scriptsize $G(150, y)$\hspace*{3em}}; &
            \\[-0.08cm]
            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize Satisfiable?}; &

@@

            \\[-0.08cm]
            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize Glasgow}; &

@@

            \\[-0.08cm]
            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize LAD}; &

@@

            \\[-0.08cm]
            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize VF2}; &

@@

            \\[-0.08cm]
            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize VF3}; &

@@

            \\[-0.08cm]
            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize Constrainedness}; &

@@

            \\[-0.08cm]
            \node [anchor=center, inner sep=2pt, rotate=90] {\scriptsize Complement?}; &

@@

    \caption{Behaviour of algorithms on the induced variant with target graphs of 150 vertices,
    shown in the style of \cref{figure:non-induced}. The sixth row plots constrainedness using
    equation~\eqref{equation:constrainedness}: the darkest region is where $\kappa = 1$, and the
    lighter regions show where the problem is either over- or under-constrained. The final row shows
    when the Glasgow algorithm performs better when given the complements of the pattern and target
    graphs as inputs---the solid lines show the location of the phase transition, and the dotted
    lines are $t_d=0.5$ and the $p_d=t_d$ diagonal.}\label{figure:induced}

@@

            \node [anchor=center] {\scriptsize $G(10, x) \hookrightarrow$\hspace*{1em}}; &
